Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-05 image_analysis_paper_development2/manuscript.tex
updated manuscript

@@

\renewcommand{\figurename}{Fig.}
\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
\newcommand{\fig}[0]{Fig.}

@@

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1],
Ioannis~G.~Kevrekidis \footnotemark[1] \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA} ,
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \footnotemark[2], \\ and
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}

@@

Imaging studies provide unique insights into the dynamics of pattern formation and morphogenesis in developing tissues.
%
Developmental progress is commonly reconstructed from snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific developmental system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
To illustrate our approach, we register and order data from imaging studies of pattern formation and morphogenesis in three different model systems in developmental biology.

\vspace{1em}

@@

We present a robust algorithmic approach to simultaneous registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of spatial landmarks for registration or markers of developmental progression.
%
The approach is based on a manifold learning algorithm (vector diffusion maps \citep{singer2012vector}) which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.
%
Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology. 
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent. 
%
Our final data set consists of z-stacks of wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data. 
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus here only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing procedures, our algorithms can also be used simply to extract the main direction of variability within the imaging data set. 
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parameterization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parameterization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parameterization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the available software. 
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parameterization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}). 
%
Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{ 
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\ 
2 & 4.98$^{\circ}$ & 0.9994 \\ 
3 & 6.47$^{\circ}$ & 0.9795 \\ 
4 & 7.03$^{\circ}$ & 0.9953 \\ 
5 & 3.04$^{\circ}$ & 0.9959 \\ 
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

To assess the robustness of our proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period. 
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E. 
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.

In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered. 
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images. 
%% 
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.  
%
In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a flip book movie of zebrafish epiboly. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the rather slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

To illustrate the broad applicability of our approach, we also applied our algorithm to a time-lapse movie  of zebrafish embryogenesis. 
%
We used a publicly available flipbook movie of zebrafish embryogenesis, available at \texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov} \citep{karlstrom1996flipbook}.  
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage. 
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization). 
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk. 
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development. 
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.  
%
During this stage, the cells rearrange to form a cell layer with uniform thickness (about four cells thick). 
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole. 
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A). 
%
We then used vector diffusion maps to register and order the frames. 
%
The results are shown in \fig~\ref{fig:zebrafish}B. 
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C). 
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$. 
%
Some errors in ordering images of the early embryo result from rather slow cell movement during the early developmental stage where cells divide and accumulate above the yolk. 
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we have showed that our algorithm is not limited to a specific set of data. 
%
The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

@@

\begin{figure*}[t]
\begin{minipage}[b]{12cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo. 
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such noise sources,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability. 
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

@@

@@

\caption{Analysis of three-dimensional wing disc data. (A) Maximum projections of an example three-dimensional z-stack. Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability. 
%
Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample. 
%
However, many developmental processes contain interesting dynamics in all three dimensions.
%
With advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.

Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computational prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle). 
%
However, ordering data via diffusion maps is readily possible for registered three-dimensional data. 
%
Therefore, we can apply our methodology to three-dimensional data by first registering the data using an alternative approach, and then ordering the registered data using diffusion maps.

To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.  
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into a specific organ, such as wings, legs, and eyes, in the adult fly during the pupal transformation. 
%
The wing disc is an imaginal disc that turns into a fly wing. 
%
The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization. 
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals. 
%
The discs were stained with antibodies that recognize Spalt (red), Dad-GFP (green), and Wingless/Patched (gray) (\fig~\ref{fig:wing_disc}A). 
%
The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

>>>>

\usepackage{color}

@@

\renewcommand{\figurename}{Fig.}
%\newcommand{\SI}[0]{\textit{SI Materials and Methods}}
\newcommand{\SI}[0]{supplementary material}
\newcommand{\fig}[0]{Fig.}

@@

\author{Carmeline~J.~Dsilva \thanks{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA},
Bomyi~Lim \footnotemark[1],
Hang~Lu \thanks{School of Chemical and Biomolecular Engineering, Georgia Institute of Technology, Atlanta, Georgia, USA},
Amit~Singer \thanks{Department of Mathematics, Princeton University, Princeton, New Jersey, USA} \thanks{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}, Ioannis~G.~Kevrekidis \footnotemark[1] \footnotemark[4] ,\\ and
Stanislav~Y.~Shvartsman \footnotemark[1] \thanks{Lewis-Sigler Institute for Integrative Genomics, Princeton University, Princeton, New Jersey, USA}}

@@

Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide a software implementation of our methodology to aid in the application of our techniques to other experimental data sets. 
\vspace{1em}

@@

We present a robust algorithmic approach to simultaneous registration and temporal ordering.
%
In contrast to a number of previous methodologies \citep{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face, dubuis2013accurate}, our methodology does not rely on the {\em a priori} knowledge of landmarks for registration or markers of developmental progression.
%
The approach is based on vector diffusion maps \citep{singer2012vector}, a manifold learning algorithm which simultaneously addresses the problems of registration and temporal ordering.
%
This algorithm is one of several nonlinear dimensionality reduction techniques that have been developed over the past decade \citep{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}, for
applications ranging from analysis of cryo-electron microscopy (cryo-EM) images of individual molecules  \citep{zhao2014rotationally, singer2011viewing} to face recognition \citep{lafon2006data} and classification of CT scans \citep{fernandez2014diffusion}.

Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology.
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
%
Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translational and/or scaling symmetries via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parameterization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

\raisebox{4.25cm}{\textsf{C}}
\includegraphics[width=5cm]{drosophila_live_imaging_angle_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{D}}
\includegraphics[width=5cm]{drosophila_live_imaging_rank_corr}
%
%\hfill
%
\raisebox{4.25cm}{\textsf{E}}
%
\raisebox{2.5cm}{
{ \textsf{
\begin{tabular}{ |c | c| c| }
\hline
\parbox[c]{1.5cm}{ \centering Movie Index} & \parbox[c]{1.5cm}{ \vspace{.5\baselineskip} \centering Average Angle Error \vspace{.5\baselineskip}} & \parbox[c]{1.5cm}{ \centering Rank Correlation Coefficient} \\
\hline
1 & 8.37$^{\circ}$ & 0.9989 \\
2 & 4.98$^{\circ}$ & 0.9994 \\
3 & 6.47$^{\circ}$ & 0.9795 \\
4 & 7.03$^{\circ}$ & 0.9953 \\
5 & 3.04$^{\circ}$ & 0.9959 \\
 \hline
\end{tabular}}}}
\caption{Method validation using live imaging of \textit{Drosophila} embryos. (A) Selected images from a live imaging study of a \textit{Drosophila} embryo during gastrulation. Scale bar indicates 50$\mu$m. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A registered and ordered by vector diffusion maps. The dorsal side of each embryo now appears at the top of each image, and the ventral side appears at the bottom. (C) The correlation between the recovered rotation angle (using vector diffusion maps) and the true rotation angle. The average absolute error in the recovered angles is $8.37 ^\circ$. (D) The correlation between the recovered rank (using vector diffusion maps) and the true rank. The rank correlation coefficient is $0.9989$. (E) The average error in the recovered angle and the rank correlation coefficient for $5$ independent live imaging studies. }
\label{fig:drosophila_live_imaging}
\end{figure*}

@@

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\vspace{0cm}
\raisebox{2.5cm}{\textsf{C}}
%
\includegraphics[width=3.25cm]{zebrafish_rank_corr}
\end{minipage}
\caption{Method validation using live imaging of a zebrafish embryo. (A) Selected images from a movie of zebrafish epiboly. Each frame is in an arbitrary rotational orientation, and the order of the frames has been shuffled. (B) Images from A after registration and ordering using vector diffusion maps. The real time of each frame is also indicated. (C) Correlation between the rank recovered using vector diffusion maps and the true rank.  The rank correlation coefficient is $0.9954$. The larger errors in the recovered ranks towards the beginning of the trajectory are due to the slow cell movement within that time window. }
\label{fig:zebrafish}
\end{figure*}

As another validation for the proposed methodology, we applied our algorithm to a time-lapse movie  of zebrafish embryogenesis.
%
We used a publicly available live imaging data of zebrafish embryogenesis (\texttt{https://zfin.org/zf\_info/movies/Zebrafish.mov}, \citet{karlstrom1996flipbook}).
%
Taken with a differential interference contrast (DIC) microscope, the movie records the first $17$~hours of zebrafish development, from a single cell stage to a 16-somite stage.
%
We selected $120$ consecutive frames from this movie which capture 5.5~hours of epiboly (3.5--9~hours after fertilization).
%
At the start of the time window, cells have divided 10--11 times and are accumulated in a cell mass above the yolk.
%
The cell mass is then compressed and the animal-vegetal axis of the embryo (vertical axis in \fig~\ref{fig:zebrafish}) shortens to form a spherical embryo shape by the end of the fourth hour of development.
%
Then, the yolk syncytial layer, which forms the boundary between the yolk and the cell mass, moves upward, forming a dome-shaped structure.
%
During this stage, the cells rearrange to form a uniform layer about four cells thick.
%
With time, this cell layer then spreads over across the yolk and expands toward the vegetal pole.
%
At the end of epiboly, the blastoderm completely engulfs the yolk.

As in the example of {\em Drosophila} embryo live imaging, the frames were randomly rotated and shuffled (\fig~\ref{fig:zebrafish}A).
%
We then used vector diffusion maps to register and order the frames.
%
The results are shown in \fig~\ref{fig:zebrafish}B.
%
The recovered rotations and order are consistent with the expected developmental dynamics, as shown in the correlations between the recovered and true ranks (\fig~\ref{fig:zebrafish}C).
%
Quantitatively, the rank correlation coefficient for this data set is 0.9954, and the average error in the recovered angle is $4.14^\circ$.
%
Some errors in ordering images of the early embryo result from slow cell movement during the early developmental stage where cells divide and accumulate above the yolk.
%
During epiboly, cell movement is more dynamic and the recovered ordering is more consistent with the real dynamics.

In summary, we have shown that our approach to temporal ordering performs very well on imaging data of two different developmental processes ({\em Drosophila} gastrulation and zebrafish epiboly), taken with two different imaging methods (fluorescent microscopy and DIC) where the true temporal order is known {\em a priori}.
%image registration and ordering can correctly recover image orientation and  showed that our algorithm is not limited to a specific set of data.
%
%The algorithm works for different tissue types ({\em Drosophila} and zebrafish), and also for images taken with different imaging methods (fluorescent and phase contrast images).
%
Provided there exist significant dynamics within the data set and that the developmental trajectory is well-sampled, the developmental dynamics can be recovered.

@@

We have analyzed how our algorithm performs on two model data sets where all images come from a single embryo.
%
In practice, we are interested in cases where image comes from a different embryo, and the largest source of noise in the considered data set arises from embryo-to-embryo variability.
%
To demonstrate that our methods are robust to such variations,
we constructed a synthetic time course data set by selecting a random image from one of five {\em Drosophila} live imaging data sets (those data sets used in \fig~\ref{fig:drosophila_live_imaging}) at each time point.
%
The resulting data set is spatially unregistered, scrambled in time, and reflects embryo-to-embryo variability.
%
The median rank correlation coefficient when ordering such a synthetic time course using our methodology was 0.77, indicating that the algorithm can recover the temporal order even under noisy conditions.

@@

\begin{figure*}[t!]
\begin{minipage}[b]{12cm}
\raisebox{0.5cm}{\textsf{A}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_scrambled}

\raisebox{0.5cm}{\textsf{C}}
\includegraphics[width=12.25cm]{drosophila_fixed_images_average}
\end{minipage}
%
\hfill
%
\raisebox{3.25cm}{\textsf{D}}
\includegraphics[width=4.25cm]{drosophila_fixed_images_rank_corr}
%
\caption{Analysis of images of fixed \textit{Drosophila} embryos. {(A)} Images of \textit{ Drosophila} embryos, stained for nuclei (gray), Twi (green), and dpERK (red). Scale bar indicates approximately 50$\mu$m (images have been rescaled to remove slight interembryo size variations).  Each image is of a different embryo arrested at a different developmental time and in a different rotational orientation. {(B)} Data from {\it A}, registered and ordered using vector diffusion maps. The expert rank for each image is indicated. {(C)} A representative ``developmental trajectory'' obtained from local averaging of the entire set of registered and ordered images (see \SI). {(D)} Correlation between the image ranks calculated from the vector diffusion maps algorithm and the ranks obtained from ordering by an expert. The rank correlation coefficient is $0.9716$. }
\label{fig:drosophila_fixed_images}
\end{figure*}

@@

\caption{Analysis of three-dimensional {\em Drosophila} wing disc z-stacks. (A) Maximum projections of an example three-dimensional {\em Drosophila} wing disc z-stack. Discs express the Dad-GFP reporter construct (green), and are stained for Spalt (red), Wingless (gray), and Patched (gray). Projections along the $x$-, $y$-, and $z$-axes are shown. (B) Example three-dimensional images, ordered using diffusion maps. The time cohort, as assessed by an expert, is indicated for each image, and the rank correlation coefficient between the diffusion maps ordering and the expert timing is 0.9427. (C) The average developmental trajectory for the registered and ordered images. }
\label{fig:wing_disc}
\end{figure*}

%We have demonstrated that our algorithm can recover the underlying developmental dynamics in several different types of imaging data sets, including ones containing interembryo variability.
%
%Thus far, we have only looked at two-dimensional data sets where biological samples are imaged as a single confocal slice or as a phase contrast image at a fixed position within the sample.
%
%However, with advances in imaging techniques, taking multiple z-slices to visualize the three-dimensional structure of a sample will become more common, and it is important to validate that our algorithm can be applied to such data.

Although the theory of vector diffusion maps readily extends to three-dimensional data with full three-dimensional rotations, calculation of the pairwise alignments rapidly becomes computational prohibitive for such applications, as one must now search over all three-dimensional rotations (which are parameterized by three Euler angles), rather than over two-dimensional rotations (which are parameterized by a {\em single} rotation angle).
%
However, ordering data via diffusion maps is readily possible for registered three-dimensional data.
%
Therefore, we can apply our methodology to three-dimensional data, such as a stack of confocal images taken at different depths within the sample, by first registering the data using an alternative method, and then ordering the registered data using diffusion maps.
